```{r 01-project-setup-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Project Setup and Reproducible Workflows

This chapter introduces the recommended project structure for this course and explains how to keep your work organized, reproducible, and easy to share. A clear project setup is one of the most important parts of data management because it helps you avoid broken file paths, lost outputs, duplicate scripts, and unclear analysis steps.

By the end of this chapter, you should be able to:

- create and use an RStudio Project;
- organize raw data, scripts, outputs, and reports into clear folders;
- use relative file paths instead of computer-specific paths;
- understand the role of Git and GitHub in reproducible analysis; and
- apply a simple workflow for saving, documenting, and sharing your work.

---

## Why project setup matters

A data analysis project usually includes several types of files, including raw data, cleaned data, R Markdown files, figures, tables, notes, and final reports. If these files are not organized consistently, the analysis becomes difficult to reproduce and difficult for others to review.

A good project structure helps you:

- find files quickly;
- avoid overwriting important work;
- keep raw data separate from cleaned or modified data;
- make your code easier to run on another computer;
- document each step of the analysis; and
- create a clear audit trail of your work.

In this course, we will use **RStudio Projects** and **R Markdown** to support a clean and reproducible workflow.

---

## Recommended folder structure

For this course, use one main project folder. Inside that folder, create separate folders for raw data, generated outputs, and any supporting materials.

A recommended structure is:

```text
Data-Management-with-R/
│
├── data_management.Rproj
├── index.Rmd
├── 01_project_setup.Rmd
├── 02_tidyverse_basics.Rmd
├── 03_joining_data.Rmd
├── 04_data_cleaning.Rmd
├── 05_strings_regex.Rmd
├── 06_visualization.Rmd
├── 07_tidyverse_analysis.Rmd
├── 08_storyboard.Rmd
│
├── Raw Data/
├── Outputs/
├── docs/
├── archive/
│
├── _bookdown.yml
├── _output.yml
├── style.css
├── README.md
├── LICENSE
└── .gitignore
```

### Folder descriptions

| Folder or file | Purpose |
|---|---|
| `data_management.Rproj` | Opens the project in RStudio and sets the working directory correctly. |
| `index.Rmd` | The landing page or introduction for the bookdown project. |
| `01_*.Rmd`, `02_*.Rmd`, etc. | Main chapter files for the book. |
| `Raw Data/` | Original input datasets. These should not be manually edited. |
| `Outputs/` | Cleaned datasets, tables, figures, and other generated files. |
| `docs/` | Rendered bookdown website output. |
| `archive/` | Old drafts, unused files, or previous versions kept for reference. |
| `_bookdown.yml` | Controls bookdown file order and output folder. |
| `_output.yml` | Controls bookdown output format. |
| `README.md` | Explains the project for users and collaborators. |
| `.gitignore` | Tells Git which files or folders should not be tracked. |

---

## Using RStudio Projects

An **RStudio Project** keeps all files for one analysis or course in a single working environment. This is better than opening individual files from different folders because RStudio knows where the project starts.

To create a project:

1. Open RStudio.
2. Select **File > New Project**.
3. Choose either **New Directory** or **Existing Directory**.
4. Save the project in the main course folder.
5. Open the `.Rproj` file whenever you work on the project.

When you open the `.Rproj` file, RStudio automatically sets the project folder as the working directory.

---

## Working with relative paths

Avoid using full file paths that only work on your computer, such as:

```{r 01-project-setup-bad_path_example, eval=FALSE}
read_csv("C:/Users/YourName/Desktop/Data/Population_Estimates.csv")
```

Instead, use relative paths from the project folder:

```{r 01-project-setup-good_path_example, eval=FALSE}
read_csv("./Raw Data/Population_Estimates.csv")
```

This makes your code easier to share and easier to run on another computer.

You can also use the `here` package:

```{r 01-project-setup-here_example, eval=FALSE}
library(here)
read_csv(here("Raw Data", "Population_Estimates.csv"))
```

Both approaches are acceptable. The key idea is to avoid paths that only work on one machine.

---

## File naming conventions

Good file names make a project easier to navigate. Use names that are short, descriptive, and consistent.

Recommended practices:

- Use lowercase where possible.
- Use underscores instead of spaces.
- Number chapter files in the order they appear.
- Avoid vague names such as `final_final_version2.Rmd`.
- Use dates only when they are truly useful.

Examples:

```text
01_project_setup.Rmd
02_tidyverse_basics.Rmd
03_joining_data.Rmd
cancer_rates.csv
exposures.csv
```

---

## Separating raw data and outputs

Raw data should stay unchanged. If you clean, transform, or summarize the data, save the result in the `Outputs/` folder instead of overwriting the original file.

For example:

```{r 01-project-setup-output_example, eval=FALSE}
rates_clean <- rates %>%
  filter(letter == "C")

write_csv(rates_clean, "./Outputs/Cancer_rates.csv")
```

This makes it easier to trace how the cleaned data was created.

---

## R Markdown workflow

R Markdown allows you to combine text, code, and results in one document. This is useful for teaching, reporting, and reproducible analysis.

A typical R Markdown workflow is:

1. Explain the purpose of the section.
2. Load the required packages.
3. Read the data.
4. Clean or reshape the data.
5. Create summaries, tables, or plots.
6. Explain what the results mean.
7. Save any outputs that are needed later.

This structure helps readers understand not only what the code does, but also why each step is needed.

---

## Bookdown chapter organization

In this course, each major topic should be placed in its own `.Rmd` file. The order of the chapters is controlled by `_bookdown.yml`.

A recommended order is:

```text
index.Rmd
01_project_setup.Rmd
02_tidyverse_basics.Rmd
03_joining_data.Rmd
04_data_cleaning.Rmd
05_strings_regex.Rmd
06_visualization.Rmd
07_tidyverse_analysis.Rmd
08_storyboard.Rmd
```

This structure keeps the book organized and makes it easier to revise one chapter without affecting the others.

---

## GitHub and version control

Git and GitHub are recommended for advanced users and for anyone who wants to track changes carefully over time.

To get the most out of this workshop, we recommend working within **RStudio Projects** (`.Rproj`) and using **Git** for version control, with **GitHub** as the remote repository.

Using Git and GitHub allows you to:

- track changes to your work over time;
- revert to earlier versions if needed;
- collaborate safely with others;
- keep a clear audit trail of your analysis; and
- back up your project outside your local computer.

RStudio includes built-in Git tools, so you can use RStudio as your Git interface without installing a separate Git graphical interface, unless you already have one that you prefer.

Throughout this workshop, the recommended workflow is:

- one RStudio Project per analysis or assignment;
- Git enabled at the project level; and
- GitHub used as the remote backup and collaboration platform.

For detailed R-focused guidance, the online book **Happy Git with R** is a useful reference. The steps below are based on that workflow.

---

## Installation and first-time Git setup

Complete these steps once on your computer.

### 1. Create a GitHub account

Create a free GitHub account:

<https://happygitwithr.com/github-acct.html>

### 2. Install Git

Git is a system-level tool that must be installed before RStudio can use it.

Guidance:

<https://happygitwithr.com/install-git.html>

Notes:

- Windows users should install **Git for Windows**.
- macOS users can use Homebrew or the Xcode Command Line Tools.

### 3. Introduce yourself to Git

Git needs your name and email address for the commit history.

Guidance:

<https://happygitwithr.com/hello-git.html>

This step connects your future commits to your identity as the project author or contributor.

---

## Connecting Git, GitHub, and RStudio

After Git is installed, connect it to GitHub and RStudio.

### 1. Connect local Git to GitHub

Guidance:

<https://happygitwithr.com/push-pull-github.html>

This allows your local project to communicate with the remote GitHub repository.

### 2. Choose an authentication method

GitHub no longer allows password-based authentication for Git operations. You must use one of the following methods.

#### Option A: HTTPS and Personal Access Token

This is recommended for most users.

<https://happygitwithr.com/credential-caching.html>

#### Option B: SSH keys

This is recommended if you work with Git frequently or across multiple computers.

<https://happygitwithr.com/ssh-keys.html>

### 3. Enable Git support inside RStudio

Guidance:

<https://happygitwithr.com/rstudio-git-github.html>

Once Git is connected, RStudio will show a **Git** tab in the project interface.

---

## Recommended workflow: GitHub first

A **GitHub-first workflow** means the repository is created on GitHub first and then cloned into RStudio.

Full walkthrough:

<https://happygitwithr.com/new-github-first.html>

### Core steps

1. Create a new repository on GitHub.
2. Clone the repository into a new RStudio Project.
3. Add or edit files locally.
4. Commit the changes in RStudio.
5. Push the committed changes back to GitHub.

This workflow is recommended because it makes the connection between GitHub and the local RStudio Project clearer from the beginning.

---

## Practical notes for workshops

Git is useful, but it is not mandatory for completing the workshop materials.

If you prefer not to use GitHub, you can still:

- work locally with `.Rproj` files;
- organize your data and outputs using the recommended folder structure;
- knit R Markdown documents; and
- build the bookdown project locally.

If you get stuck with Git or GitHub, continue working locally and sync later when the issue is resolved.

---

## Optional GitHub extensions

After you are comfortable with the basics, you may explore:

- GitHub Desktop: <https://desktop.github.com/>
- branches;
- pull requests;
- issues;
- README-driven documentation; and
- GitHub Pages for publishing project websites.

---

## Chapter summary

In this chapter, you learned how to structure a reproducible R project. The key recommendations are:

- use one RStudio Project per analysis or course;
- keep raw data in `Raw Data/`;
- save generated files in `Outputs/`;
- use relative paths;
- organize bookdown chapters with numbered `.Rmd` files;
- use Git for version tracking when appropriate; and
- use GitHub for backup and collaboration when needed.

A clean project structure makes your analysis easier to understand, easier to reproduce, and easier to maintain.
