# Tidyverse Basics

The **tidyverse** is a collection of R packages designed for data science and data analysis. These packages provide a consistent and readable approach for importing, cleaning, transforming, visualizing, and analyzing data. In practice, the tidyverse is widely used because it allows complex workflows to be written in a clear and organized way. The tidyverse ecosystem is widely used for data science workflows @wickham2023.

In this chapter, you will learn how to load tidyverse packages, read datasets into R, inspect data frames, and perform common data manipulation tasks using `dplyr`. By the end of the chapter, you should feel comfortable performing basic exploratory data analysis and creating simple data transformation pipelines.


## Loading packages and reading data

Before working with data in R, the required packages must first be loaded. The `tidyverse` package includes several commonly used packages such as `dplyr`, `ggplot2`, `readr`, and `tidyr`.

```{r 02-tidyverse-basics-load-tidyverse}
library(tidyverse)
```

In most projects, the first analytical step is reading raw data into R. Throughout this course, raw datasets should be stored inside the `Raw Data/` folder introduced in the previous chapter.

The example below demonstrates how to read a `.csv` file using `read_csv()` from the `readr` package.

```{r 02-tidyverse-basics-read-example-data, eval=FALSE}
example_data <- read_csv(
  "./Raw Data/example.csv"
)
```

Using relative paths helps keep projects reproducible because the code can run correctly on different computers as long as the project structure remains consistent.

::: {.tip}
One of the most common beginner problems in R involves incorrect working directories or broken file paths. Using RStudio Projects and relative paths helps avoid many of these issues.
:::

## Inspecting data frames

After importing a dataset, it is important to understand its structure before beginning any analysis. This step is often called *data inspection* or *exploratory inspection*.

Several functions are useful for quickly examining a data frame.

```{r 02-tidyverse-basics-inspect-functions, eval=FALSE}
glimpse(example_data)

summary(example_data)

dim(example_data)

names(example_data)
```

The `glimpse()` function provides a compact overview of the dataset structure, including variable names and data types. The `summary()` function generates summary statistics for each variable, while `dim()` returns the number of rows and columns in the dataset. The `names()` function lists all variable names.

Inspecting data early in the workflow helps identify potential issues such as missing values, incorrect variable types, or unexpected column names.

## Data manipulation with dplyr

One of the most powerful features of the tidyverse is the `dplyr` package, which provides functions for manipulating and transforming data frames.

Data manipulation workflows are commonly built using the pipe operator `%>%`. The pipe sends the output of one step directly into the next step, making code easier to read and understand.

The example below demonstrates several commonly used `dplyr` verbs.

```{r 02-tidyverse-basics-dplyr-verbs-example, eval=FALSE}
example_data %>%
  
  select(column_1, column_2) %>%
  
  filter(column_1 == "value") %>%
  
  mutate(new_column = column_2 * 2) %>%
  
  group_by(column_1) %>%
  
  summarize(
    total = sum(new_column, na.rm = TRUE)
  )
```

In this workflow:

- `select()` chooses specific columns from the dataset.
- `filter()` keeps only rows that satisfy a condition.
- `mutate()` creates a new variable.
- `group_by()` creates groups within the data.
- `summarize()` calculates summary statistics for each group.

Together, these functions form the foundation of many data analysis workflows in R.

::: {.note}
The pipe operator `%>%` can be interpreted as “then.” Reading pipelines from top to bottom often makes the analysis logic easier to follow.
:::

## Practice exercise

To reinforce the concepts introduced in this chapter, try completing the following short exercise using one of your own datasets or a practice dataset provided in the course materials.

1. Import a `.csv` dataset from the `Raw Data/` folder.
2. Use `glimpse()` and `summary()` to inspect the dataset.
3. Select several variables of interest.
4. Filter the data to one category or group.
5. Create a new variable using `mutate()`.
6. Calculate a simple summary statistic using `summarize()`.

As you practice these steps, focus not only on writing code that works, but also on creating workflows that are readable, organized, and reproducible.

## Chapter summary

In this chapter, you were introduced to the tidyverse and several core tools used in modern R workflows. You learned how to load packages, import datasets, inspect data frames, and manipulate data using common `dplyr` functions.

These foundational skills will be used throughout the remainder of the course for data cleaning, visualization, exploratory analysis, and reporting.
