# Tidyverse Basics

## Learning objectives

By the end of this chapter, you should be able to:

- load common R packages;
- read data files into R;
- inspect data frames;
- use the pipe operator;
- apply common `dplyr` verbs such as `select`, `filter`, `mutate`, `group_by`, and `summarize`.

## Load libraries

```{r 02-tidyverse-basics-load-tidyverse}
library(tidyverse)
```

## Read data

In most projects, the first step is reading raw data from the `Raw Data/` folder.

```{r 02-tidyverse-basics-read-example-data, eval=FALSE}
example_data <- read_csv("./Raw Data/example.csv")
```

## Inspect data

Useful functions include:

```{r 02-tidyverse-basics-inspect-functions, eval=FALSE}
glimpse(example_data)
summary(example_data)
dim(example_data)
names(example_data)
```

## Core tidyverse verbs

```{r 02-tidyverse-basics-dplyr-verbs-example, eval=FALSE}
example_data %>%
  select(column_1, column_2) %>%
  filter(column_1 == "value") %>%
  mutate(new_column = column_2 * 2) %>%
  group_by(column_1) %>%
  summarize(total = sum(new_column, na.rm = TRUE))
```

## Practice questions

1. Read one `.csv` file from your `Raw Data/` folder.
2. Use `glimpse()` to inspect the dataset.
3. Select three columns.
4. Filter the data to one group or category.
5. Create one new variable with `mutate()`.
