# Joining Data

## Learning objectives

By the end of this chapter, you should be able to:

- identify keys used to join datasets;
- understand one-to-one, one-to-many, and many-to-many joins;
- use `left_join()`, `inner_join()`, and related join functions;
- check for duplicate keys before joining.

## Example data

```{r 03-joining-data-join-example-data}
library(tidyverse)

x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     1, "x4"
)

y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2"
)
```

## Left join

A left join keeps all rows from the first dataset and brings in matching information from the second dataset.

```{r 03-joining-data-left-join-example}
left_join(x, y, by = "key")
```

## Check duplicate keys

Before joining real datasets, always check whether the join key is unique.

```{r 03-joining-data-check-duplicate-keys}
x %>% count(key) %>% filter(n > 1)
y %>% count(key) %>% filter(n > 1)
```

## Common join types

- `left_join()`: keep all rows from the left table.
- `inner_join()`: keep only matched rows.
- `right_join()`: keep all rows from the right table.
- `full_join()`: keep all rows from both tables.
- `anti_join()`: find rows that do not have a match.

## Practice

1. Create two small datasets with a common key.
2. Try `left_join()` and `inner_join()`.
3. Check what happens when one table has duplicate keys.
4. Use `anti_join()` to identify unmatched rows.
