# Data Cleaning and Data Management

Data cleaning is one of the most important stages of any analytical workflow. In real-world projects, raw data are rarely ready for analysis immediately after import. Datasets often contain inconsistent variable names, missing values, formatting issues, duplicate records, encoding problems, or variables stored in unsuitable formats.

This chapter introduces several common data management tasks using the tidyverse. You will learn how to import multiple file types, inspect datasets, identify common data quality issues, reshape data into tidy formats, and recode variables into more useful forms.

Although the datasets used in this course are simplified for teaching purposes, the workflow closely reflects many real-world data cleaning tasks encountered in epidemiology, environmental health, and population data analysis.

## Preparing the R environment

Before beginning any analysis, the required libraries should be loaded. For this chapter, we will primarily use packages from the tidyverse ecosystem, along with `ggplot2` and `readxl`.

If these packages are not already installed on your computer, they can be installed using:

```{r 04-data-cleaning-install, eval=FALSE}
install.packages(
  c("tidyverse", "ggplot2", "readxl")
)
```

Once installed, the libraries can be loaded into the R session.

```{r 04-data-cleaning-load-libraries}
library(tidyverse)
library(ggplot2)
library(readxl)
```

## Exploring project files

Before importing data, it is often useful to inspect the contents of the project folders. The `list.files()` function allows you to display all files stored within a directory.

```{r 04-data-cleaning-list-files}
list.files("./Raw Data/")
```

You can also search for specific file types using the `pattern` argument.

```{r 04-data-cleaning-list-files-type}
list.files(
  "./Raw Data/",
  pattern = ".csv"
)

list.files(
  "./Raw Data/",
  pattern = ".xlsx"
)
```

Functions such as `list.files()` are frequently overlooked by beginners, but they become extremely useful when working with large projects that contain many datasets or outputs.

Some datasets may also be stored in compressed `.zip` files. These files should typically be extracted before analysis.

```{r 04-data-cleaning-unzip, eval=FALSE}
unzip(
  "./Raw Data/Corr_2016.zip",
  exdir = "./Raw Data"
)
```

The `exdir` argument specifies the folder where the extracted files should be saved. Since files only need to be unzipped once, this code is often commented out after the extraction is complete.

::: {.tip}
When working in R Markdown documents, avoid repeatedly running file extraction code unless necessary. Re-extracting files every time the document is knitted can slow down workflows and create duplicate files.
:::

## Description of the datasets

This course uses four datasets that simulate a simplified population health analysis workflow.

The mortality dataset contains records of deaths, including demographic information and ICD-10 cause-of-death codes. Although the dataset is based on realistic structures and counts, the data themselves are fictitious for teaching purposes.

The population dataset contains estimates of population counts by age group, sex, and Health Service Delivery Area in British Columbia.

The correspondence dataset links lower geographic units to larger health regions and service delivery areas. Correspondence files are commonly used in population health and environmental research because they allow different geographic datasets to be connected.

The environmental dataset contains weather-related variables by postal code, including measures related to temperature and cold exposure.

The overall analytical goal is to combine these datasets in order to calculate mortality rates and explore potential environmental relationships.

## Importing data into R

Different file formats require different import functions. The `readr` package is commonly used for `.csv` files, while Excel files require functions from the `readxl` package.

The following code imports the datasets used in this chapter.

```{r 04-data-cleaning-load-data}
mort <- read_excel(
  "./Raw Data/deaths_2016.xlsx"
)

pop <- read_csv(
  "./Raw Data/Population_Estimates.csv"
)

corr <- read_csv(
  "./Raw Data/Corr_2016.csv",
  locale = readr::locale(
    encoding = "latin1"
  )
)

env <- read_csv(
  "./Raw Data/Weather_data.csv"
)
```

Notice that the correspondence file uses a specific text encoding. Encoding problems are common when datasets contain accented characters or multilingual text.

## Inspecting datasets

After importing data, it is important to examine the structure and contents of each dataset before beginning analysis.

Several functions provide useful high-level summaries.

```{r 04-data-cleaning-EDA}
str(mort)
summary(mort)

dplyr::glimpse(pop, width = 60)
summary(pop)

dplyr::glimpse(corr, width = 60)
summary(corr)

str(env)
summary(env)
```

The `str()` function displays the structure of the dataset, including variable types and dimensions. The `summary()` function provides summary statistics and quick descriptive information.

Another useful function is `glimpse()` from `dplyr`.

```{r 04-data-cleaning-glimpse}
glimpse(mort)
```

Compared with `str()`, `glimpse()` often provides a cleaner and easier-to-read overview of the data frame.

Initial data inspection is one of the most important parts of data cleaning because it helps identify:

- missing values;
- unusual variable names;
- incorrect data types;
- inconsistent formatting; and
- unexpected values.

## Exploring and validating the data

A useful next step is performing simple exploratory checks to understand the datasets more fully.

For example, we can identify all health regions in the correspondence dataset.

```{r 04-data-cleaning-health-regions}
corr %>%
  distinct(hrname_english) %>%
  mutate(
    hrname_english = iconv(
      hrname_english,
      from = "latin1",
      to = "UTF-8",
      sub = ""
    )
  ) %>%
  arrange(hrname_english)
```

When working with multilingual datasets, encoding issues may cause characters to display incorrectly. Re-importing the file using the appropriate encoding often resolves these problems.

```{r 04-data-cleaning-reload-corr}
corr <- read_csv(
  "./Raw Data/Corr_2016.csv",
  locale = readr::locale(
    encoding = "latin1"
  )
)

corr %>%
  distinct(hrname_english) %>%
  arrange(hrname_english)
```

We can also calculate the number of health regions in the dataset.

```{r 04-data-cleaning-number-of-regions}
no_regions <- corr %>%
  distinct(hrname_english) %>%
  nrow()
```

There are `r no_regions` health regions in the correspondence dataset.

Additional exploratory summaries can help validate the imported data.

```{r 04-data-cleaning-investigation}
mort %>%
  count(Sex)

pop %>%
  summarize(
    Total_below_1 = sum(`<1`)
  )
```

These simple checks help verify that the datasets were imported correctly and that the variables contain plausible values.

## Cleaning variable names and reshaping data

One common issue in real-world datasets is poorly formatted variable names. Variables that begin with numbers or contain symbols can make coding more difficult.

For example:

```{r 04-data-cleaning-column-names}
names(pop)

colnames(pop)
```

Some age-group variables contain names such as `"04-Jan"` or `"09-May"`, which are not ideal variable names for analysis.The `janitor` package is also useful for cleaning messy column names and inspecting datasets @janitor.

One approach is to rename problematic variables directly.

```{r 04-data-cleaning-rename-columns}
pop2 <- pop %>%
  rename(
    "01-04" = "04-Jan",
    "05-09" = "09-May",
    "10-14" = "14-Oct"
  )
```

Another important data-cleaning task is reshaping data into a *tidy* format. In tidy data, each variable has its own column, each observation has its own row, and each value occupies a single cell.

The `pivot_longer()` function is commonly used to convert wide datasets into long format.

```{r 04-data-cleaning-pivot}
pop_long <- pop %>%
  
  select(
    -any_of(c("X1", "Total"))
  ) %>%
  
  pivot_longer(
    cols = -any_of(
      c(
        "Year",
        "Gender",
        "Health Service Delivery Area"
      )
    ),
    names_to = "Age",
    values_to = "Value"
  )

pop_long
```

The resulting dataset stores age groups within a single variable rather than across many separate columns.

After reshaping the data, some values may still require cleaning or recoding. The `case_when()` function is often used for this purpose.

```{r 04-data-cleaning-reformat}
pop_long <- pop_long %>%
  
  mutate(
    Age = case_when(
      Age == "04-Jan" ~ "01-04",
      Age == "09-May" ~ "05-09",
      Age == "14-Oct" ~ "10-14",
      TRUE ~ Age
    )
  )
```

The final line, `TRUE ~ Age`, is important because it preserves all values that do not match the earlier conditions.

If this line is removed, unmatched rows become missing values.

```{r 04-data-cleaning-recode-wrong}
pop %>%
  
  select(
    -any_of(c("X1", "Total"))
  ) %>%
  
  pivot_longer(
    cols = -any_of(
      c(
        "Year",
        "Gender",
        "Health Service Delivery Area"
      )
    ),
    names_to = "Age",
    values_to = "Value"
  ) %>%
  
  mutate(
    Age = case_when(
      Age == "04-Jan" ~ "01-04",
      Age == "09-May" ~ "05-09",
      Age == "14-Oct" ~ "10-14"
    )
  )
```

::: {.warning}
Always verify the output after recoding variables. Small coding mistakes can unintentionally convert valid values into missing values.
:::

## Practical considerations for R Markdown workflows

When working with R Markdown documents, remember that every code chunk is executed again during knitting. Code that modifies files, unzips folders, or downloads data should therefore be used carefully.

As projects grow larger, keeping code organized and avoiding repeated operations becomes increasingly important for reproducible workflows.

You should also begin thinking about how analyses are documented. A good R Markdown document not only produces correct outputs, but also explains the reasoning behind each analytical step.

## Chapter summary

In this chapter, you learned several important data cleaning and management techniques using the tidyverse. You imported multiple datasets, explored project files, inspected dataset structures, handled encoding issues, cleaned variable names, reshaped data into tidy formats, and recoded variables using `case_when()`.

These skills form the foundation of reproducible data analysis workflows and will be used extensively throughout the remainder of the course.
