# Strings and Regular Expressions

Character variables are extremely common in real-world datasets. Names, addresses, postal codes, diagnostic codes, survey responses, and file names are all examples of text-based data that often require cleaning before analysis can begin.

This chapter introduces tools for working with strings and regular expressions in R using the tidyverse and the `stringr` package. You will learn how to separate text variables into multiple columns, remove unnecessary spaces, extract patterns from text, reshape complex datasets, and use regular expressions to identify meaningful information hidden inside character strings. The `stringr` package simplifies many common string operations in R @stringr. 

In addition to text cleaning, this chapter also introduces important data reshaping techniques using `pivot_longer()`, which is commonly used to convert messy wide-format datasets into tidy long-format structures suitable for analysis.

## Working with character data in R

The tidyverse includes several useful tools for working with text data. In this chapter, we will primarily use functions from `stringr`, which provides a consistent and readable framework for string manipulation.

Begin by loading the required libraries.

```{r 05-strings-regex-load-libraries}
library(tidyverse)
library(stringr)
```

Throughout this chapter, we will work with two example datasets:

- the Titanic dataset, which will be used to practice cleaning names and extracting passenger titles; and
- the WHO tuberculosis dataset, which will be used to demonstrate reshaping and decoding complex variable names.

## Cleaning and separating character variables

The Titanic dataset contains a variable called `Name`, which stores multiple pieces of information within a single text field.

We begin by importing the data and converting selected variables into factors.

```{r 05-strings-regex-read-titanic-data}
titanic <- read_csv(
  "./Raw Data/titanic.csv",
  col_types = cols(
    Survived = col_factor(),
    Sex = col_factor()
  )
)
```

Before working with the names, we perform a small amount of initial cleaning. Records with missing survival information are removed, and a new variable called `family_size` is created.

```{r 05-strings-regex-clean-titanic-data}
titanic <- titanic %>%
  
  filter(!is.na(Survived)) %>%
  
  mutate(
    family_size = SibSp + Parch + 1
  )

head(titanic)
```

The `Name` variable contains both last names and first names in a single column. To make the data easier to analyze, we can separate this variable into two new columns.

```{r 05-strings-regex-separate-name}
titanic <- titanic %>%
  
  separate(
    col = Name,
    into = c("Last_name", "First_name"),
    sep = ","
  )

head(
  titanic %>%
    select(Last_name, First_name)
)
```

After separating the names, some records contain extra spaces at the beginning of the `First_name` variable. These spaces can be removed using `str_trim()`.

```{r 05-strings-regex-trim-first-name}
titanic <- titanic %>%
  
  mutate(
    First_name = str_trim(
      First_name,
      side = "left"
    )
  )

head(
  titanic %>%
    select(Last_name, First_name)
)
```

Cleaning whitespace is a small but important part of text processing because extra spaces can interfere with matching, grouping, and filtering operations.

## Extracting patterns with regular expressions

Many text variables contain structured patterns that can be extracted using regular expressions, often called *regex*.

In the Titanic dataset, passenger titles such as `"Mr"`, `"Mrs"`, `"Miss"`, and `"Master"` appear within the `First_name` variable. We can extract these titles using the `str_extract()` function.

```{r 05-strings-regex-extract-title}
titanic <- titanic %>%
  
  mutate(
    Title = str_extract(
      First_name,
      "^[^.]+"
    )
  )

head(
  titanic %>%
    select(First_name, Title)
)
```

The regular expression `^[^.]+` can be interpreted as follows:

- `^` indicates the beginning of the string;
- `[^.]` means any character except a period; and
- `+` means one or more occurrences of the previous pattern.

Together, the expression extracts all characters from the beginning of the string until the first period is reached.

Regular expressions can initially appear intimidating, but they become extremely powerful once you begin recognizing common text patterns.

::: {.tip}
When learning regular expressions, it is often helpful to interpret them one symbol at a time rather than trying to understand the entire pattern at once.
:::

## Creating reusable functions

When the same sequence of operations must be repeated multiple times, it is often useful to place the code inside a function.

The following function trims whitespace and extracts passenger titles.

```{r 05-strings-regex-create-title-function}
create_title <- function(firstname) {
  
  firstname %>%
    
    str_trim(side = "left") %>%
    
    str_extract("^[^.]+")
}
```

We can then apply the function directly to the dataset.

```{r 05-strings-regex-test-title-function}
titanic <- titanic %>%
  
  mutate(
    Title_from_function =
      create_title(First_name)
  )

head(
  titanic %>%
    select(
      First_name,
      Title,
      Title_from_function
    )
)
```

Creating functions improves readability and reduces repeated code, especially in large projects.

## Summarizing and visualizing categorical data

Once titles have been extracted, we can summarize the number of passengers within each title group.

```{r 05-strings-regex-title-summary}
title_summary <- titanic %>%
  
  count(Title, sort = TRUE)

title_summary
```

We can also visualize the distribution using `ggplot2`.

```{r 05-strings-regex-plot-title-summary}
title_summary %>%
  
  ggplot(
    aes(
      x = reorder(Title, n),
      y = n
    )
  ) +
  
  geom_col() +
  
  coord_flip() +
  
  labs(
    title = "Number of Titanic Passengers by Title",
    x = "Title",
    y = "Number of Passengers"
  )
```

Visual summaries are often useful for identifying unusual categories, inconsistencies, or unexpected values within character variables.

## Reshaping complex datasets

The WHO tuberculosis dataset provides an excellent example of why data reshaping is often necessary.

```{r 05-strings-regex-load-who-data}
data(who)

glimpse(who)
```

Many column names in this dataset are not true variables. Instead, the column names themselves contain several pieces of information.

For example:

```text
new_sp_m014
```

contains:

- the diagnosis type;
- the sex category; and
- the age group.

This structure violates the principles of tidy data because multiple variables are embedded inside the column names.

To clean the dataset, we first reshape it from wide format into long format using `pivot_longer()`.

```{r 05-strings-regex-pivot-who-longer}
who1 <- who %>%
  
  pivot_longer(
    cols = new_sp_m014:newrel_f65,
    names_to = "key",
    values_to = "cases",
    values_drop_na = TRUE
  )

who1
```

After reshaping the data, we correct inconsistent variable naming.

```{r 05-strings-regex-clean-who-key}
who2 <- who1 %>%
  
  mutate(
    key = str_replace(
      key,
      "newrel",
      "new_rel"
    )
  )

who2
```

The combined `key` variable can then be separated into multiple meaningful variables.

```{r 05-strings-regex-separate-who-key}
who3 <- who2 %>%
  
  separate(
    col = key,
    into = c(
      "new",
      "type",
      "sexage"
    ),
    sep = "_"
  )

who3
```

Finally, the `sexage` variable can be divided into separate sex and age variables.

```{r 05-strings-regex-separate-sex-age}
who4 <- who3 %>%
  
  separate(
    col = sexage,
    into = c("sex", "age"),
    sep = 1
  )

who4
```

The entire cleaning workflow can also be written as one continuous pipeline.

```{r 05-strings-regex-full-who-pipeline}
who_clean <- who %>%
  
  pivot_longer(
    cols = new_sp_m014:newrel_f65,
    names_to = "key",
    values_to = "cases",
    values_drop_na = TRUE
  ) %>%
  
  mutate(
    key = str_replace(
      key,
      "newrel",
      "new_rel"
    )
  ) %>%
  
  separate(
    key,
    into = c(
      "new",
      "type",
      "sexage"
    ),
    sep = "_"
  ) %>%
  
  select(
    -new,
    -iso2,
    -iso3
  ) %>%
  
  separate(
    sexage,
    into = c("sex", "age"),
    sep = 1
  )

who_clean
```

## Using regular expressions inside pivot_longer()

An alternative approach uses regular expressions directly inside `pivot_longer()` to extract multiple variables at once.

```{r 05-strings-regex-regex-practice}
who_longer <- who %>%
  
  pivot_longer(
    cols = new_sp_m014:newrel_f65,
    
    names_to = c(
      "diagnosis",
      "gender",
      "age"
    ),
    
    names_pattern =
      "new_?(.*)_(.)(.*)",
    
    values_to = "count",
    
    values_drop_na = TRUE
  )

who_longer
```

The regular expression used here contains several capture groups:

- `(.*)` extracts the diagnosis type;
- `(.)` extracts one character representing sex; and
- the final `(.*)` extracts the age group.

These extracted components are automatically assigned to the variables listed in `names_to`.

This approach demonstrates how regular expressions can be integrated directly into data reshaping workflows.

## Missing values and interpretation

During reshaping, we used the argument:

```r
values_drop_na = TRUE
```

This removes rows containing missing values.

However, it is important to remember that missing values are not always equivalent to zero values.

An `NA` typically means that the value is unknown or unavailable, while a value of `0` indicates that the quantity is known and equal to zero.

Understanding the meaning of missing values is an essential part of responsible data analysis.

::: {.warning}
Never assume that missing values and zeros represent the same thing. Incorrect handling of missing data can substantially alter analytical results.
:::

## Additional practice

The cleaned datasets created in this chapter can now be used for additional exploratory analysis.

Possible exercises include:

- identifying the most common passenger titles in the Titanic dataset;
- creating survival summaries by sex or passenger class;
- summarizing tuberculosis cases by age group or sex; and
- visualizing categorical distributions using bar plots.

These exercises help reinforce the connection between text cleaning, data reshaping, and exploratory analysis.

## Chapter summary

In this chapter, you learned how to work with character variables and regular expressions using the tidyverse and `stringr`. You cleaned and separated text variables, extracted patterns using regex, created reusable functions, reshaped wide datasets into tidy long-format structures, and interpreted coded variable names.

These techniques are extremely important in real-world data management because many datasets contain poorly structured text variables and encoded information that must be cleaned before analysis can begin.
