# Strings and Regular Expressions

```{r 05-strings-regex-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Learning Objectives

By the end of this practice, you should be able to:

- Load and inspect data using tidyverse tools.
- Separate one character column into multiple columns.
- Clean extra spaces using `str_trim()`.
- Extract text patterns using regular expressions and `str_extract()`.
- Create a simple custom function.
- Reshape data from wide format to long format using `pivot_longer()`.
- Separate coded variable names into meaningful variables.
- Summarize and visualize categorical data.

# 1. Load Libraries

We will use the `tidyverse` package, which includes `dplyr`, `tidyr`, `ggplot2`, `readr`, and other useful packages. We will also load `stringr` for working with text.

```{r 05-strings-regex-load-libraries}
library(tidyverse)
library(stringr)
```

# 2. Titanic Example: Cleaning Names and Extracting Titles

In this section, we use the Titanic dataset to practice working with character variables.

## 2.1 Read the Data

The dataset is read from the `Raw Data` folder. We also convert `Survived` and `Sex` to factor variables during import.

```{r 05-strings-regex-read-titanic-data}
titanic <- read_csv(
  "./Raw Data/titanic.csv",
  col_types = cols(
    Survived = col_factor(),
    Sex = col_factor()
  )
)
```

## 2.2 Basic Cleaning

First, we remove records where `Survived` is missing. Then, we create a new variable called `family_size`.

```{r 05-strings-regex-clean-titanic-data}
titanic <- titanic %>%
  filter(!is.na(Survived)) %>%
  mutate(family_size = SibSp + Parch + 1)

head(titanic)
```

# 3. Separate Passenger Names

The `Name` column contains both last name and first name information. We can separate this column into two columns: `Last_name` and `First_name`.

```{r 05-strings-regex-separate-name}
titanic <- titanic %>%
  separate(
    col = Name,
    into = c("Last_name", "First_name"),
    sep = ","
  )

head(titanic %>% select(Last_name, First_name))
```

## 3.1 Remove Extra Spaces

After separating the names, the `First_name` column may contain extra spaces at the beginning. We can remove these using `str_trim()`.

```{r 05-strings-regex-trim-first-name}
titanic <- titanic %>%
  mutate(First_name = str_trim(First_name, side = "left"))

head(titanic %>% select(Last_name, First_name))
```

# 4. Extract Titles Using Regex

Passenger titles such as `Mr`, `Mrs`, `Miss`, and `Master` are stored inside the `First_name` column. We can extract them using `str_extract()`.

```{r 05-strings-regex-extract-title}
titanic <- titanic %>%
  mutate(Title = str_extract(First_name, "^[^.]+"))

head(titanic %>% select(First_name, Title))
```

## 4.1 Regex Explanation

The pattern `^[^.]+` means:

- `^` starts matching at the beginning of the string.
- `[^.]` means any character except a period.
- `+` means one or more of the previous character pattern.

Together, this extracts all characters from the start of the string until the first period.

# 5. Create a Function

Now we can turn the title-extraction steps into a reusable function.

```{r 05-strings-regex-create-title-function}
create_title <- function(firstname) {
  firstname %>%
    str_trim(side = "left") %>%
    str_extract("^[^.]+")
}
```

## 5.1 Test the Function

```{r 05-strings-regex-test-title-function}
titanic <- titanic %>%
  mutate(Title_from_function = create_title(First_name))

head(titanic %>% select(First_name, Title, Title_from_function))
```

# 6. Practice Question: Summarize Titanic Titles

Create a summary table showing the number of passengers in each title group.

```{r 05-strings-regex-title-summary}
title_summary <- titanic %>%
  count(Title, sort = TRUE)

title_summary
```

## 6.1 Plot Titanic Titles

```{r 05-strings-regex-plot-title-summary}
title_summary %>%
  ggplot(aes(x = reorder(Title, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Number of Titanic Passengers by Title",
    x = "Title",
    y = "Number of Passengers"
  )
```

# 7. WHO Data Example: Reshaping Data

The `who` dataset contains tuberculosis cases by country, year, sex, age group, and diagnosis type. The dataset is included in the `tidyr` package, which is part of the tidyverse.

```{r 05-strings-regex-load-who-data}
data(who)

glimpse(who)
```

# 8. Why Reshape the WHO Data?

Many columns in the WHO dataset are not true variables. Instead, their names contain several pieces of information, such as diagnosis type, sex, and age group.

For example, a column name like `new_sp_m014` contains:

- `new`: new cases
- `sp`: smear-positive pulmonary TB
- `m`: male
- `014`: age group 0–14

To make the dataset tidy, we need to reshape it from wide format to long format.

# 9. Pivot WHO Data from Wide to Long Format

```{r 05-strings-regex-pivot-who-longer}
who1 <- who %>%
  pivot_longer(
    cols = new_sp_m014:newrel_f65,
    names_to = "key",
    values_to = "cases",
    values_drop_na = TRUE
  )

who1
```

# 10. Clean Inconsistent Column Names

Some column names use `newrel` instead of `new_rel`. We fix this before separating the `key` column.

```{r 05-strings-regex-clean-who-key}
who2 <- who1 %>%
  mutate(key = str_replace(key, "newrel", "new_rel"))

who2
```

# 11. Separate the Key Column

The `key` column contains multiple pieces of information. We separate it into `new`, `type`, and `sexage`.

```{r 05-strings-regex-separate-who-key}
who3 <- who2 %>%
  separate(
    col = key,
    into = c("new", "type", "sexage"),
    sep = "_"
  )

who3
```

# 12. Separate Sex and Age

The `sexage` column combines sex and age group. We separate it after the first character.

```{r 05-strings-regex-separate-sex-age}
who4 <- who3 %>%
  separate(
    col = sexage,
    into = c("sex", "age"),
    sep = 1
  )

who4
```

# 13. Full WHO Cleaning Pipeline

The full process can be written as one pipe.

```{r 05-strings-regex-full-who-pipeline}
who_clean <- who %>%
  pivot_longer(
    cols = new_sp_m014:newrel_f65,
    names_to = "key",
    values_to = "cases",
    values_drop_na = TRUE
  ) %>%
  mutate(key = str_replace(key, "newrel", "new_rel")) %>%
  separate(key, into = c("new", "type", "sexage"), sep = "_") %>%
  select(-new, -iso2, -iso3) %>%
  separate(sexage, into = c("sex", "age"), sep = 1)

who_clean
```

# 14. WHO Data Dictionary

The coded column names contain the following information:

| Code | Meaning |
|---|---|
| `new` | New TB cases |
| `rel` | Relapse cases |
| `ep` | Extrapulmonary TB |
| `sn` | Pulmonary TB, smear negative |
| `sp` | Pulmonary TB, smear positive |
| `m` | Male |
| `f` | Female |
| `014` | 0–14 years old |
| `1524` | 15–24 years old |
| `2534` | 25–34 years old |
| `3544` | 35–44 years old |
| `4554` | 45–54 years old |
| `5564` | 55–64 years old |
| `65` | 65 years or older |

# 15. Regex Practice Question

The following code reshapes the WHO dataset using a regular expression.

```{r 05-strings-regex-regex-practice}
who_longer <- who %>%
  pivot_longer(
    cols = new_sp_m014:newrel_f65,
    names_to = c("diagnosis", "gender", "age"),
    names_pattern = "new_?(.*)_(.)(.*)",
    values_to = "count",
    values_drop_na = TRUE
  )

who_longer
```

## 15.1 Explanation of the Regex

The pattern `new_?(.*)_(.)(.*)` means:

- `new` matches the text `new` at the beginning.
- `_?` means the underscore after `new` may or may not be present.
- `(.*)` captures the diagnosis type, such as `sp`, `ep`, `sn`, or `rel`.
- `_` matches the underscore before sex and age.
- `(.)` captures one character for gender, such as `m` or `f`.
- `(.*)` captures the remaining characters as the age group.

The brackets create capture groups. These captured parts are placed into the new columns listed in `names_to`.

# 16. Discussion Question: NA vs Zero

In this practice, we used `values_drop_na = TRUE`. This removes missing values while reshaping the dataset.

This can be reasonable when we know that missing values represent combinations that were not recorded in the dataset. However, an `NA` is not always the same as zero:

- `NA` means the value is missing or unknown.
- `0` means the value is known and equal to zero.

Before dropping missing values, it is important to understand how missing values are used in the dataset.

# 17. Additional Practice Questions

Try answering the following questions using the cleaned datasets.

## 17.1 Titanic Practice

1. Which title appears most often in the Titanic dataset?
2. How many unique titles are there?
3. Create a bar plot of survival counts by sex.
4. Create a summary table showing average family size by passenger class.

## 17.2 WHO Practice

1. How many total TB cases are recorded by sex?
2. Which country has the highest number of recorded TB cases?
3. Which age group has the highest number of recorded TB cases?
4. Create a bar plot of total TB cases by age group.

# 18. Example Solutions for Extra Practice

## 18.1 Titanic: Survival Count by Sex

```{r 05-strings-regex-titanic-survival-by-sex}
titanic %>%
  count(Sex, Survived) %>%
  ggplot(aes(x = Sex, y = n, fill = Survived)) +
  geom_col(position = "dodge") +
  labs(
    title = "Titanic Survival Count by Sex",
    x = "Sex",
    y = "Number of Passengers",
    fill = "Survived"
  )
```

## 18.2 WHO: Total Cases by Sex

```{r 05-strings-regex-who-cases-by-sex}
who_clean %>%
  group_by(sex) %>%
  summarize(total_cases = sum(cases, na.rm = TRUE)) %>%
  arrange(desc(total_cases))
```

## 18.3 WHO: Total Cases by Age Group

```{r 05-strings-regex-who-cases-by-age}
who_clean %>%
  group_by(age) %>%
  summarize(total_cases = sum(cases, na.rm = TRUE)) %>%
  arrange(desc(total_cases)) %>%
  ggplot(aes(x = reorder(age, total_cases), y = total_cases)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Total TB Cases by Age Group",
    x = "Age Group",
    y = "Total Cases"
  )
```

# 19. Key Takeaways

In this practice, you learned how to:

- Clean and separate character columns.
- Use `stringr` functions to trim and extract text.
- Use regular expressions to identify text patterns.
- Reshape data from wide format to long format.
- Convert coded column names into meaningful variables.
- Summarize and visualize categorical data.
