# Visualization and Advanced Cleaning

```{r 06-visualization-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Purpose of this practice

This practice builds on the previous data management exercises. The main goals are to:

- review data cleaning and reshaping with `tidyverse`
- create simple visualizations with `ggplot2`
- identify possible join keys across datasets
- clean mortality, population, correspondence, and environmental data
- prepare cleaned output files for later analysis

# Research questions

**Analysis 1:** Calculate age- and sex-specific cancer mortality rates by health region in BC.

**Analysis 2:** Link environmental data to cancer mortality data to perform an odds ratio analysis.

> Note: In a real-world analysis, cancer case data would usually be more appropriate than mortality data. For this course, the focus is on data cleaning and data management rather than the final interpretation of the analysis.

# 1. Load packages

```{r 06-visualization-load_packages}
library(tidyverse)
library(readxl)
library(lubridate)
library(stringr)
```

# 2. Read data

The exercise uses four datasets:

| Dataset | File | Description |
|---|---|---|
| Mortality data | `deaths_2016.xlsx` | Fictitious mortality records |
| Population data | `Population_Estimates.csv` | BC population estimates by age, sex, and HSDA |
| Correspondence file | `Corr_2016.csv` | Geographic lookup between dissemination blocks and health regions |
| Environmental data | `Weather_data.csv` | Weather variables by postal code |

```{r 06-visualization-read_data}
mort <- read_excel("./Raw Data/deaths_2016.xlsx")
pop  <- read_csv("./Raw Data/Population_Estimates.csv")
corr <- read_csv(
  "./Raw Data/Corr_2016.csv",
  locale = readr::locale(encoding = "latin1")
)
env  <- read_csv("./Raw Data/Weather_data.csv")
```

# 3. Initial data review

Before cleaning, it is helpful to review the dimensions, names, missing values, and structure of each dataset.

```{r 06-visualization-create_data_list}
data_list <- list(
  mort = mort,
  pop = pop,
  corr = corr,
  env = env
)
```

## 3.1 Dataset dimensions

```{r 06-visualization-dataset_dimensions}
map_dfr(
  data_list,
  ~ tibble(rows = nrow(.x), columns = ncol(.x)),
  .id = "dataset"
)
```

## 3.2 Column names

```{r 06-visualization-column_names}
map(data_list, names)
```

## 3.3 Missing values by dataset

```{r 06-visualization-missing_values_summary}
map(
  data_list,
  ~ map_int(.x, ~ sum(is.na(.x)))
)
```

# 4. Clean and reshape population data

The population data is currently in a wide format, with age groups stored as separate columns. For plotting and analysis, it is easier to work with tidy long-format data.

## 4.1 Convert population data to long format

```{r 06-visualization-population_long}
pop_long <- pop %>%
  select(-any_of(c("X1", "Total"))) %>%
  pivot_longer(
    cols = -any_of(c("Year", "Gender", "Health Service Delivery Area")),
    names_to = "Age",
    values_to = "Population"
  ) %>%
  mutate(
    Age = case_when(
      Age == "04-Jan" ~ "01-04",
      Age == "09-May" ~ "05-09",
      Age == "14-Oct" ~ "10-14",
      TRUE ~ Age
    )
  )

pop_long
```

## 4.2 Rename columns and remove unnecessary fields

The column name `Health Service Delivery Area` is long and contains spaces. We will rename it to `HSDA`. We will also remove the `Year` column because all records are from 2016.

```{r 06-visualization-clean_population_names}
pop_long <- pop_long %>%
  rename(HSDA = `Health Service Delivery Area`) %>%
  select(-Year)

pop_long
```

## 4.3 Keep total population only

The `Gender` column includes `T`, which represents total population. Because we are interested in total population by HSDA and age group for this example, we will keep only these records.

```{r 06-visualization-filter_population_total}
pop_total <- pop_long %>%
  filter(Gender == "T")

pop_total
```

# 5. Visualize population data

## 5.1 Line plot by HSDA

A line plot can quickly show how population changes across age groups, but it is not always the best choice because age groups are categories rather than continuous values.

```{r 06-visualization-population_line_plot}
pop_total %>%
  ggplot(aes(x = Age, y = Population, group = HSDA, colour = HSDA)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Population by Age Group and HSDA",
    x = "Age group",
    y = "Population",
    colour = "HSDA"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5.2 Bar plot by age group

For reporting, a bar graph may be easier to interpret because age groups are categorical.

```{r 06-visualization-population_bar_plot}
pop_total %>%
  ggplot(aes(x = Age, y = Population, fill = HSDA)) +
  geom_col(position = "dodge") +
  labs(
    title = "Population Distribution by Age Group and HSDA",
    x = "Age group",
    y = "Population",
    fill = "HSDA"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5.3 Optional: Faceted visualization

Facets are useful when the plot becomes too busy.

```{r 06-visualization-population_facets}
pop_total %>%
  ggplot(aes(x = Age, y = Population)) +
  geom_col() +
  facet_wrap(~ HSDA) +
  labs(
    title = "Population by Age Group for Each HSDA",
    x = "Age group",
    y = "Population"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 6. Identify possible join keys

A join requires one or more common variables between datasets. Before joining, always check whether the keys are clean and whether the relationship is one-to-one, one-to-many, or many-to-many.

```{r 06-visualization-review_join_keys}
map(data_list, names)
```

Possible join paths:

- `mort` can be joined to `corr` using `dbuid2016`.
- `pop` can potentially be joined to `corr` using health region or health service delivery area variables, but the names and codes need checking first.
- `env` contains postal code data, but the link to mortality records needs postal code cleaning and validation.

# 7. Clean mortality data

## 7.1 Remove columns not needed for the current analysis

```{r 06-visualization-create_rates_data}
rates <- mort %>%
  select(-c(Location_of_death, Marital_status))

rates
```

## 7.2 Check and remove duplicate records

```{r 06-visualization-check_duplicates}
rates_dup <- rates %>%
  count(ID) %>%
  filter(n > 1)

rates_dup
```

```{r 06-visualization-inspect_duplicate_records}
rates %>%
  filter(ID %in% rates_dup$ID)
```

```{r 06-visualization-remove_duplicates}
rates <- rates %>%
  distinct()
```

# 8. Work with missing and unexpected values

Missing values can appear in different ways, such as `NA`, blanks, `NULL`, `999`, or other placeholder values. Always check how missing or unusual data is represented before making cleaning decisions.

## 8.1 Check birth year values

```{r 06-visualization-check_birth_year}
rates %>%
  ggplot(aes(x = B_year)) +
  geom_histogram(binwidth = 5) +
  labs(
    title = "Distribution of Birth Year",
    x = "Birth year",
    y = "Count"
  ) +
  theme_minimal()
```

Birth years before 1905 are unlikely in this dataset. For this exercise, we assume these values are data-entry errors where `18xx` should be `19xx`.

```{r 06-visualization-clean_birth_values}
rates <- rates %>%
  mutate(
    B_year = if_else(B_year < 1905, B_year + 100, B_year),
    B_month = if_else(is.na(B_month), 6, B_month),
    B_day = if_else(is.na(B_day), 15, B_day)
  )
```

## 8.2 Check death month and death day

```{r 06-visualization-check_death_month}
rates %>% count(D_month)
```

```{r 06-visualization-clean_death_month}
rates <- rates %>%
  mutate(D_month = if_else(D_month > 12, 6, D_month))

rates %>% count(D_month)
```

```{r 06-visualization-check_days}
rates %>% count(D_day)
rates %>% count(B_day)
```

# 9. Create date variables and calculate age

We will use `lubridate` to create date of birth (`DOB`) and date of death (`DOD`). Then we calculate age as the difference between year of death and year of birth.

```{r 06-visualization-create_dates}
rates <- rates %>%
  mutate(
    DOB = as_date(str_c(B_year, B_month, B_day, sep = "-")),
    DOD = as_date(str_c(D_year, D_month, D_day, sep = "-")),
    Age = year(DOD) - year(DOB),
    Age = if_else(Age < 0, 0.5, as.numeric(Age))
  )

range(rates$Age, na.rm = TRUE)
```

## 9.1 Check remaining missing values

```{r 06-visualization-check_remaining_na}
map_int(rates, ~ sum(is.na(.x)))
```

```{r 06-visualization-records_with_any_na}
rates %>%
  filter(if_any(everything(), is.na))
```

## 9.2 Drop records missing essential identifiers

For this exercise, records missing `B_year` or `dbuid2016` are removed because these values are required for age calculation and geographic linkage.

```{r 06-visualization-drop_essential_na}
rates <- rates %>%
  drop_na(B_year, dbuid2016)
```

## 9.3 Impute missing sex values

For practice purposes, missing `Sex` values are randomly assigned as `M` or `F`. The `set.seed()` function makes the random assignment reproducible.

```{r 06-visualization-impute_sex}
set.seed(123)

# Count and locate missing Sex values.
missing_sex_index <- is.na(rates$Sex)
missing_sex_count <- sum(missing_sex_index)

# Replace only the missing rows. This avoids the recycling error that can happen
# when using if_else() with a shorter sampled vector.
rates$Sex <- as.character(rates$Sex)
rates$Sex[missing_sex_index] <- sample(
  c("M", "F"),
  size = missing_sex_count,
  replace = TRUE
)
rates$Sex <- as.factor(rates$Sex)

rates %>% count(Sex)
```

## 9.4 Recalculate age where needed

Some records may still have missing age values because day and month values may have been swapped. For this exercise, we try an alternate date construction.

```{r 06-visualization-fix_remaining_age}
rates <- rates %>%
  mutate(
    DOB = if_else(
      is.na(DOB),
      as_date(str_c(B_year, B_day, B_month, sep = "-")),
      DOB
    ),
    Age = year(DOD) - year(DOB),
    Age = if_else(Age < 0, 0.5, as.numeric(Age))
  )

rates %>% filter(is.na(Age))
```

# 10. Work with ICD-10 cause-of-death codes

Cause-of-death codes that start with `C` represent malignant neoplasms. Codes beginning with `D` include in situ, benign, or unknown-status neoplasms and are not counted as cancer deaths in this exercise.

## 10.1 Separate ICD-10 letters and numbers

```{r 06-visualization-separate_icd_codes}
rates <- rates %>%
  extract(
    Cause_of_death,
    into = c("letter", "number"),
    regex = "([A-Z]+)([0-9]+)",
    remove = FALSE
  ) %>%
  mutate(number = as.numeric(number))

rates %>% count(letter)
```

## 10.2 Keep cancer deaths only

```{r 06-visualization-filter_cancer_codes}
rates <- rates %>%
  filter(letter == "C")
```

# 11. Categorize cancer type and age group

Cancer categories used in this practice:

| Category | ICD-10 code(s) | Description |
|---|---|---|
| 1 | C50 | Breast cancer |
| 2 | C61 | Prostate cancer |
| 3 | C33, C34 | Lung cancer |
| 4 | C18-C21 | Colorectal cancer |
| 5 | All other C codes | Other cancers |

```{r 06-visualization-categorize_cancer_and_age}
rates <- rates %>%
  mutate(
    CauseCat = case_when(
      number == 50 ~ 1,
      number == 61 ~ 2,
      number %in% c(33, 34) ~ 3,
      number %in% 18:21 ~ 4,
      TRUE ~ 5
    ),
    CauseCat_label = case_when(
      CauseCat == 1 ~ "Breast cancer",
      CauseCat == 2 ~ "Prostate cancer",
      CauseCat == 3 ~ "Lung cancer",
      CauseCat == 4 ~ "Colorectal cancer",
      CauseCat == 5 ~ "Other cancer"
    ),
    AgeCat = cut(
      Age,
      breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 130),
      include.lowest = TRUE,
      right = FALSE
    )
  )

rates %>% count(CauseCat_label)
rates %>% count(AgeCat)
```

# 12. Visualize cleaned mortality data

```{r 06-visualization-cancer_category_plot}
rates %>%
  count(CauseCat_label) %>%
  ggplot(aes(x = reorder(CauseCat_label, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Cancer Deaths by Cancer Category",
    x = "Cancer category",
    y = "Number of deaths"
  ) +
  theme_minimal()
```

```{r 06-visualization-age_category_plot}
rates %>%
  count(AgeCat, Sex) %>%
  ggplot(aes(x = AgeCat, y = n, fill = Sex)) +
  geom_col(position = "dodge") +
  labs(
    title = "Cancer Deaths by Age Group and Sex",
    x = "Age group",
    y = "Number of deaths",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 13. Save cleaned cancer mortality data

Create the `Outputs` folder if it does not already exist.

```{r 06-visualization-create_outputs_folder}
dir.create("./Outputs", showWarnings = FALSE)
```

```{r 06-visualization-save_cancer_rates}
write_csv(rates, "./Outputs/Cancer_rates.csv")
```

# 14. Clean environmental exposure data

For Analysis 2, we need to identify whether each postal code is located in a cold climate. A cold climate is defined as:

- average daily high temperature less than 10°C, and
- average daily low temperature less than -3°C.

The required variables are:

| Original variable | New variable | Meaning |
|---|---|---|
| `POSTALCODE12` | `Postal_code` | Postal code |
| `WTHNRC12_04` | `max_temp` | Annual average of daily maximum temperature |
| `WTHNRC12_05` | `min_temp` | Annual average of daily minimum temperature |

```{r 06-visualization-clean_environmental_data}
env_clean <- env %>%
  select(POSTALCODE12, WTHNRC12_04, WTHNRC12_05) %>%
  rename(
    Postal_code = POSTALCODE12,
    max_temp = WTHNRC12_04,
    min_temp = WTHNRC12_05
  ) %>%
  mutate(
    exposure = if_else(max_temp < 10 & min_temp < -3, 1, 0)
  )

env_clean
```

```{r 06-visualization-exposure_summary}
env_clean %>% count(exposure)
```

## 14.1 Save exposure data

```{r 06-visualization-save_exposure_data}
write_csv(env_clean, "./Outputs/exposures.csv")
```

# 15. Join mortality data to correspondence file

For Analysis 1, the cleaned cancer mortality data needs to be joined to the correspondence file using `dbuid2016`.

A dissemination block (`DB`) is a geographic area identifier. Many people can live in one dissemination block, but each dissemination block should correspond to one health region.

```{r 06-visualization-join_rates_corr}
analysis1 <- inner_join(rates, corr, by = "dbuid2016")

analysis1
```

## 15.1 Save joined Analysis 1 data

```{r 06-visualization-save_analysis1}
write_csv(analysis1, "./Outputs/analysis1.csv")
```

# 16. Postal code quality check activity

Postal codes often require cleaning before they can be used for joins. In this activity, we examine whether postal codes match the expected Canadian postal code format.

## 16.1 Standardize postal codes

```{r 06-visualization-standardize_postal_codes}
rates <- rates %>%
  mutate(
    Postalcode_clean = Postalcode %>%
      str_to_upper() %>%
      str_replace_all("\\s+", "")
  )
```

## 16.2 Identify postal codes that do not match the expected format

```{r 06-visualization-check_postal_regex}
postal_pattern <- "^(?!.*[DFIOQU])[A-VXY][0-9][A-Z][0-9][A-Z][0-9]$"

postal_issues <- rates %>%
  filter(is.na(str_match(Postalcode_clean, postal_pattern))) %>%
  select(ID, Postalcode, Postalcode_clean)

postal_issues
```

## 16.3 Check each postal code position

```{r 06-visualization-postal_code_positions}
rates %>%
  transmute(
    position_1 = str_sub(Postalcode_clean, 1, 1),
    position_2 = str_sub(Postalcode_clean, 2, 2),
    position_3 = str_sub(Postalcode_clean, 3, 3),
    position_4 = str_sub(Postalcode_clean, 4, 4),
    position_5 = str_sub(Postalcode_clean, 5, 5),
    position_6 = str_sub(Postalcode_clean, 6, 6)
  ) %>%
  map(unique)
```

## 16.4 Check postal code length

```{r 06-visualization-postal_code_length}
rates %>%
  mutate(postal_length = str_length(Postalcode_clean)) %>%
  count(postal_length)
```

# 17. Practice activity: Join population and correspondence data

The population and correspondence datasets do not have a perfectly clean join key. You may need to clean text or numeric identifiers before joining.

Possible approaches:

- clean and compare HSDA names in `pop` and `corr`
- compare code-based fields such as `X1` and `hruid2017`
- reshape the population data before joining
- check how many records match and how many do not

```{r 06-visualization-practice_join_pop_corr, eval=FALSE}
# Example starting point only.
# You will need to inspect the exact values before deciding on the best join.

pop_join_ready <- pop_total %>%
  mutate(HSDA_clean = str_to_lower(HSDA) %>% str_squish())

corr_join_ready <- corr %>%
  mutate(hrname_clean = str_to_lower(hrname_english) %>% str_squish())

pop_corr_test <- left_join(
  pop_join_ready,
  corr_join_ready,
  by = c("HSDA_clean" = "hrname_clean")
)

pop_corr_test
```

# 18. Final assignment guidance

Create a new R Markdown document for your first exploratory data analysis file. Choose a clear file name, such as:

- `EDA_01.Rmd`
- `CancerMortality_01.Rmd`
- `Assignment1.Rmd`
- `exploratory_analysis.Rmd`

Your document should explain the data and cleaning process clearly enough for a colleague or manager to follow.

Include:

- a list of your datasets
- a function that can be applied to your dataset list using `purrr`
- a summary of dataset dimensions
- a description of what each column represents
- high-level statistics for each dataset
- missing value counts and how missing values were handled
- plots that help explain the data
- a clean section for population data cleaning and joining
- clear comments explaining what each major code block does and why

# 19. Reflection questions

1. Why is long-format data usually easier to plot with `ggplot2`?
2. Why should we check join keys before merging datasets?
3. What is the difference between an `NA` value and a true zero?
4. Why is it important to document assumptions when cleaning dates?
5. What are the risks of randomly imputing missing values?
6. Why might postal codes need standardization before joining datasets?
