# Visualization and Advanced Data Cleaning

In previous chapters, we focused on importing, cleaning, reshaping, and joining datasets using the tidyverse. In this chapter, we extend those workflows by introducing exploratory visualization and more advanced data-cleaning techniques commonly used in population health and environmental data analysis.

The overall objective is to prepare several datasets for later analytical work involving cancer mortality, population estimates, geographic correspondence files, and environmental exposure data. Along the way, we will review common challenges such as missing values, inconsistent variable names, duplicate records, invalid dates, and poorly formatted identifiers.

Although the datasets used in this course are simplified for teaching purposes, the workflow closely reflects the types of cleaning and preparation steps encountered in real-world epidemiological and environmental health projects.

The chapter also introduces several important visualization techniques using `ggplot2`, which will help you explore and validate your cleaned datasets.

## Research context and datasets

The exercises in this chapter are motivated by two simplified research questions.

The first analysis investigates age- and sex-specific cancer mortality rates by health region in British Columbia. The second analysis explores whether environmental exposure variables can be linked to mortality data in order to perform a simplified odds-ratio analysis.

In practice, incidence or cancer case data would generally be preferable to mortality data for exposure studies. However, the emphasis of this course is on data cleaning and management rather than formal epidemiological interpretation.

Four datasets will be used throughout the chapter:

- a mortality dataset containing fictitious death records;
- a population dataset containing age- and sex-specific population estimates;
- a geographic correspondence file linking dissemination blocks to health regions; and
- an environmental dataset containing weather variables by postal code.

We begin by loading the required libraries.

```{r 06-visualization-load-packages}
library(tidyverse)
library(readxl)
library(lubridate)
library(stringr)
```

Next, we import the datasets into R.

```{r 06-visualization-read-data}
mort <- read_excel(
  "./Raw Data/deaths_2016.xlsx"
)

pop <- read_csv(
  "./Raw Data/Population_Estimates.csv"
)

corr <- read_csv(
  "./Raw Data/Corr_2016.csv",
  locale = readr::locale(
    encoding = "latin1"
  )
)

env <- read_csv(
  "./Raw Data/Weather_data.csv"
)
```

## Reviewing datasets before cleaning

Before modifying any dataset, it is important to understand its overall structure. Initial exploratory review often helps identify unexpected variable names, missing values, duplicate records, or formatting inconsistencies.

We first create a list containing all datasets.

```{r 06-visualization-create-data-list}
data_list <- list(
  mort = mort,
  pop = pop,
  corr = corr,
  env = env
)
```

We can then quickly review dataset dimensions.

```{r 06-visualization-dataset-dimensions}
map_dfr(
  data_list,
  ~ tibble(
    rows = nrow(.x),
    columns = ncol(.x)
  ),
  .id = "dataset"
)
```

Column names can also be inspected.

```{r 06-visualization-column-names}
map(data_list, names)
```

Finally, we summarize missing values across datasets.

```{r 06-visualization-missing-values-summary}
map(
  data_list,
  ~ map_int(.x, ~ sum(is.na(.x)))
)
```

Early exploratory review is one of the most important stages of data cleaning because it helps identify potential problems before they propagate through the analysis workflow.

::: {.tip}
Always inspect datasets immediately after import. Many analytical problems originate from unnoticed import issues, incorrect variable types, or unexpected missing values.
:::

## Cleaning and reshaping population data

The population dataset is currently stored in a wide format, where age groups appear as separate columns. Although this structure may be useful for spreadsheets, it is less convenient for visualization and analysis.

A tidy long-format structure is generally easier to work with in R.

We first reshape the dataset using `pivot_longer()`.

```{r 06-visualization-population-long}
pop_long <- pop %>%
  
  select(
    -any_of(c("X1", "Total"))
  ) %>%
  
  pivot_longer(
    cols = -any_of(
      c(
        "Year",
        "Gender",
        "Health Service Delivery Area"
      )
    ),
    names_to = "Age",
    values_to = "Population"
  ) %>%
  
  mutate(
    Age = case_when(
      Age == "04-Jan" ~ "01-04",
      Age == "09-May" ~ "05-09",
      Age == "14-Oct" ~ "10-14",
      TRUE ~ Age
    )
  )

pop_long
```

Some variable names are unnecessarily long or contain spaces, so we simplify them.

```{r 06-visualization-clean-population-names}
pop_long <- pop_long %>%
  
  rename(
    HSDA = `Health Service Delivery Area`
  ) %>%
  
  select(-Year)

pop_long
```

For this example, we focus on total population counts only.

```{r 06-visualization-filter-population-total}
pop_total <- pop_long %>%
  
  filter(Gender == "T")

pop_total
```

## Visualizing population data

Visualization is an important part of exploratory analysis because it helps reveal trends, inconsistencies, and unusual values that may not be obvious in tables alone.

The following line plot displays population counts across age groups by HSDA.

```{r 06-visualization-population-line-plot}
pop_total %>%
  
  ggplot(
    aes(
      x = Age,
      y = Population,
      group = HSDA,
      colour = HSDA
    )
  ) +
  
  geom_line() +
  
  geom_point() +
  
  labs(
    title = "Population by Age Group and HSDA",
    x = "Age group",
    y = "Population",
    colour = "HSDA"
  ) +
  
  theme_minimal() +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

Because age groups are categorical rather than continuous, a bar plot may sometimes be easier to interpret.

```{r 06-visualization-population-bar-plot}
pop_total %>%
  
  ggplot(
    aes(
      x = Age,
      y = Population,
      fill = HSDA
    )
  ) +
  
  geom_col(position = "dodge") +
  
  labs(
    title = "Population Distribution by Age Group and HSDA",
    x = "Age group",
    y = "Population",
    fill = "HSDA"
  ) +
  
  theme_minimal() +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

When plots become crowded, faceting can improve readability.

```{r 06-visualization-population-facets}
pop_total %>%
  
  ggplot(
    aes(
      x = Age,
      y = Population
    )
  ) +
  
  geom_col() +
  
  facet_wrap(~ HSDA) +
  
  labs(
    title = "Population by Age Group for Each HSDA",
    x = "Age group",
    y = "Population"
  ) +
  
  theme_minimal() +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

## Identifying join keys

Before combining datasets, it is important to identify variables that can serve as join keys.

```{r 06-visualization-review-join-keys}
map(data_list, names)
```

Possible join paths include:

- joining mortality data to the correspondence file using `dbuid2016`;
- joining population data to geographic information using HSDA-related variables; and
- linking environmental data to mortality records using postal codes.

Before performing joins, always inspect whether the key variables are unique and consistently formatted.

::: {.warning}
Incorrect joins are one of the most common causes of duplicate records and analytical errors in real-world projects.
:::

## Cleaning mortality data

We now begin cleaning the mortality dataset.

Variables not required for the current analysis are removed.

```{r 06-visualization-create-rates-data}
rates <- mort %>%
  
  select(
    -c(
      Location_of_death,
      Marital_status
    )
  )

rates
```

Duplicate records should also be identified and reviewed.

```{r 06-visualization-check-duplicates}
rates_dup <- rates %>%
  
  count(ID) %>%
  
  filter(n > 1)

rates_dup
```

```{r 06-visualization-inspect-duplicate-records}
rates %>%
  
  filter(ID %in% rates_dup$ID)
```

After inspection, duplicates can be removed.

```{r 06-visualization-remove-duplicates}
rates <- rates %>%
  
  distinct()
```

## Handling missing and inconsistent values

Missing values and data-entry problems are common in administrative datasets.

We begin by examining birth year distributions.

```{r 06-visualization-check-birth-year}
rates %>%
  
  ggplot(
    aes(x = B_year)
  ) +
  
  geom_histogram(binwidth = 5) +
  
  labs(
    title = "Distribution of Birth Year",
    x = "Birth year",
    y = "Count"
  ) +
  
  theme_minimal()
```

Some birth years before 1905 are likely data-entry errors. For this exercise, we assume these values should belong to the twentieth century.

```{r 06-visualization-clean-birth-values}
rates <- rates %>%
  
  mutate(
    B_year = if_else(
      B_year < 1905,
      B_year + 100,
      B_year
    ),
    
    B_month = if_else(
      is.na(B_month),
      6,
      B_month
    ),
    
    B_day = if_else(
      is.na(B_day),
      15,
      B_day
    )
  )
```

We also inspect death months and days.

```{r 06-visualization-check-death-month}
rates %>%
  count(D_month)
```

```{r 06-visualization-clean-death-month}
rates <- rates %>%
  
  mutate(
    D_month = if_else(
      D_month > 12,
      6,
      D_month
    )
  )

rates %>%
  count(D_month)
```

```{r 06-visualization-check-days}
rates %>%
  count(D_day)

rates %>%
  count(B_day)
```

## Creating dates and calculating age

The `lubridate` package simplifies date handling in R.

We create dates of birth and death and then calculate age.

```{r 06-visualization-create-dates}
rates <- rates %>%
  
  mutate(
    DOB = as_date(
      str_c(
        B_year,
        B_month,
        B_day,
        sep = "-"
      )
    ),
    
    DOD = as_date(
      str_c(
        D_year,
        D_month,
        D_day,
        sep = "-"
      )
    ),
    
    Age = year(DOD) - year(DOB),
    
    Age = if_else(
      Age < 0,
      0.5,
      as.numeric(Age)
    )
  )

range(rates$Age, na.rm = TRUE)
```

Remaining missing values can then be inspected.

```{r 06-visualization-check-remaining-na}
map_int(
  rates,
  ~ sum(is.na(.x))
)
```

```{r 06-visualization-records-with-any-na}
rates %>%
  
  filter(
    if_any(
      everything(),
      is.na
    )
  )
```

Records missing essential identifiers are removed.

```{r 06-visualization-drop-essential-na}
rates <- rates %>%
  
  drop_na(
    B_year,
    dbuid2016
  )
```

## Imputing missing values

For demonstration purposes, missing sex values are randomly imputed.

```{r 06-visualization-impute-sex}
set.seed(123)

missing_sex_index <- is.na(rates$Sex)

missing_sex_count <- sum(missing_sex_index)

rates$Sex <- as.character(rates$Sex)

rates$Sex[missing_sex_index] <- sample(
  c("M", "F"),
  size = missing_sex_count,
  replace = TRUE
)

rates$Sex <- as.factor(rates$Sex)

rates %>%
  count(Sex)
```

Although this approach is useful for teaching purposes, real-world imputation should be performed much more carefully.

::: {.warning}
Random imputation can introduce bias into analyses. Always document assumptions and cleaning decisions clearly.
:::

## Working with ICD-10 codes

Cause-of-death variables often contain coded medical classifications.

We separate ICD-10 codes into letter and numeric components.

```{r 06-visualization-separate-icd-codes}
rates <- rates %>%
  
  extract(
    Cause_of_death,
    into = c("letter", "number"),
    regex = "([A-Z]+)([0-9]+)",
    remove = FALSE
  ) %>%
  
  mutate(
    number = as.numeric(number)
  )

rates %>%
  count(letter)
```

Cancer-related mortality records are identified using ICD-10 codes beginning with `"C"`.

```{r 06-visualization-filter-cancer-codes}
rates <- rates %>%
  
  filter(letter == "C")
```

We then classify cancer categories and age groups.

```{r 06-visualization-categorize-cancer-and-age}
rates <- rates %>%
  
  mutate(
    CauseCat = case_when(
      number == 50 ~ 1,
      number == 61 ~ 2,
      number %in% c(33, 34) ~ 3,
      number %in% 18:21 ~ 4,
      TRUE ~ 5
    ),
    
    CauseCat_label = case_when(
      CauseCat == 1 ~ "Breast cancer",
      CauseCat == 2 ~ "Prostate cancer",
      CauseCat == 3 ~ "Lung cancer",
      CauseCat == 4 ~ "Colorectal cancer",
      CauseCat == 5 ~ "Other cancer"
    ),
    
    AgeCat = cut(
      Age,
      breaks = c(
        0, 10, 20, 30, 40,
        50, 60, 70, 80, 90, 130
      ),
      include.lowest = TRUE,
      right = FALSE
    )
  )

rates %>%
  count(CauseCat_label)

rates %>%
  count(AgeCat)
```

## Visualizing cleaned mortality data

After cleaning the mortality data, we can explore the distributions visually.

```{r 06-visualization-cancer-category-plot}
rates %>%
  
  count(CauseCat_label) %>%
  
  ggplot(
    aes(
      x = reorder(CauseCat_label, n),
      y = n
    )
  ) +
  
  geom_col() +
  
  coord_flip() +
  
  labs(
    title = "Cancer Deaths by Cancer Category",
    x = "Cancer category",
    y = "Number of deaths"
  ) +
  
  theme_minimal()
```

```{r 06-visualization-age-category-plot}
rates %>%
  
  count(AgeCat, Sex) %>%
  
  ggplot(
    aes(
      x = AgeCat,
      y = n,
      fill = Sex
    )
  ) +
  
  geom_col(position = "dodge") +
  
  labs(
    title = "Cancer Deaths by Age Group and Sex",
    x = "Age group",
    y = "Number of deaths",
    fill = "Sex"
  ) +
  
  theme_minimal() +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

## Saving cleaned datasets

Cleaned outputs should always be saved separately from raw data.

We first create an output folder if it does not already exist.

```{r 06-visualization-create-outputs-folder}
dir.create(
  "./Outputs",
  showWarnings = FALSE
)
```

We then save the cleaned mortality data.

```{r 06-visualization-save-cancer-rates}
write_csv(
  rates,
  "./Outputs/Cancer_rates.csv"
)
```

Environmental exposure data can also be cleaned and saved.

```{r 06-visualization-clean-environmental-data}
env_clean <- env %>%
  
  select(
    POSTALCODE12,
    WTHNRC12_04,
    WTHNRC12_05
  ) %>%
  
  rename(
    Postal_code = POSTALCODE12,
    max_temp = WTHNRC12_04,
    min_temp = WTHNRC12_05
  ) %>%
  
  mutate(
    exposure = if_else(
      max_temp < 10 &
      min_temp < -3,
      1,
      0
    )
  )

env_clean
```

```{r 06-visualization-save-exposure-data}
write_csv(
  env_clean,
  "./Outputs/exposures.csv"
)
```

## Joining datasets and validating postal codes

The mortality dataset can now be joined to the correspondence file.

```{r 06-visualization-join-rates-corr}
analysis1 <- inner_join(
  rates,
  corr,
  by = "dbuid2016"
)

analysis1
```

Postal codes should also be standardized before linkage.

```{r 06-visualization-standardize-postal-codes}
rates <- rates %>%
  
  mutate(
    Postalcode_clean =
      Postalcode %>%
      
      str_to_upper() %>%
      
      str_replace_all(
        "\\s+",
        ""
      )
  )
```

We can then identify postal codes that do not follow the expected Canadian format.

```{r 06-visualization-check-postal-regex}
postal_pattern <-
  "^(?!.*[DFIOQU])[A-VXY][0-9][A-Z][0-9][A-Z][0-9]$"

postal_issues <- rates %>%
  
  filter(
    is.na(
      str_match(
        Postalcode_clean,
        postal_pattern
      )
    )
  ) %>%
  
  select(
    ID,
    Postalcode,
    Postalcode_clean
  )

postal_issues
```

Regular-expression validation is an extremely useful technique for detecting formatting problems in identifiers such as postal codes, phone numbers, or IDs.

## Final assignment guidance

As you continue building analytical workflows, begin organizing your work into separate exploratory data analysis documents.

Each analysis file should clearly explain:

- the datasets being used;
- the cleaning and validation steps performed;
- assumptions made during cleaning;
- how missing values were handled; and
- how outputs were generated.

A well-documented workflow is just as important as the analysis itself.

## Chapter summary

In this chapter, you expanded your data management workflow by combining advanced cleaning techniques with exploratory visualization. You reviewed datasets, reshaped population data, created visualizations, identified join keys, cleaned mortality and environmental datasets, handled missing values, validated identifiers, and created cleaned outputs for later analysis.

These workflows form the foundation of reproducible analytical projects and closely resemble the types of preparation steps required in real-world public health and environmental data analysis.
