# Exploratory Analysis Project

## Learning objectives

By the end of this chapter, you should be able to:

- create a short exploratory analysis report;
- summarize the dimensions and structure of several datasets;
- count missing values;
- create reusable functions and apply them with `purrr::map()`;
- explain cleaning decisions in plain language.

## Purpose

This chapter turns the earlier practice work into a more polished exploratory analysis document. The goal is to create a report that a colleague or manager could read to understand the data, what was cleaned, and what decisions were made.

## Load libraries

```{r 07-exploratory-analysis-eda-libraries}
library(tidyverse)
library(readxl)
library(lubridate)
```

## Read data

```{r 07-exploratory-analysis-eda-read-data, eval=FALSE}
mort <- read_excel("./Raw Data/deaths_2016.xlsx")
pop  <- read_csv("./Raw Data/Population_Estimates.csv")
corr <- read_csv("./Raw Data/Corr_2016.csv", locale = readr::locale(encoding = "latin1"))
env  <- read_csv("./Raw Data/Weather_data.csv")
```

## Create a data list

```{r 07-exploratory-analysis-eda-data-list, eval=FALSE}
data_list <- list(
  mortality = mort,
  population = pop,
  correspondence = corr,
  environment = env
)
```

## Dataset dimensions

```{r 07-exploratory-analysis-eda-dimensions, eval=FALSE}
map(data_list, dim)
map(data_list, nrow)
map(data_list, ncol)
```

## Missing values function

```{r 07-exploratory-analysis-missing-values-function}
count_missing <- function(data) {
  data %>% summarize(across(everything(), ~sum(is.na(.))))
}
```

```{r 07-exploratory-analysis-apply-missing-function, eval=FALSE}
map(data_list, count_missing)
```

## Recommended report structure

Use this structure for your exploratory analysis R Markdown file:

1. Purpose of the analysis
2. Data sources
3. Data dimensions
4. Column descriptions
5. Missing value checks
6. Duplicate checks
7. Cleaning decisions
8. Initial plots
9. Summary of key findings
10. Next steps

## Practice task

Create a new R Markdown document called `exploratory_analysis.Rmd`. Use the structure above and include at least one table, one missing-value summary, and one plot.
