---
title: "titanic_demo_two"
author: "Noushin Nabavi"
date: '2020-01-21'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidyr)
library(stringr)
library(purrr)
```


```{r echo=TRUE}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     1, "x4"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2"
)
left_join(x, y, by = "key")

```

```{r echo=TRUE}

x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     3, "x4"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     2, "y3",
     3, "y4"
)
left_join(x, y, by = "key")
```

#Advanced Mutate

###Used to be `mutate_at` `mutate_if`, etc. In the most recent version of `dplyr` the syntax is now `across`.
across() makes it easy to apply the same transformation to multiple columns, allowing you to use select() semantics inside in summarise() and mutate()

```{r across}
# A named list of functions
iris %>%
  dplyr::group_by(Species) %>%
  dplyr::summarise(across(starts_with("Sepal"), list(mean = mean, sd = sd)))

```

```{across2}
iris %>%
  group_by(Species) %>%
  summarise(across(starts_with("Sepal"), mean, .names = "mean_{col}"))

```

# Functions

Need to load data below for an output: (See exercise)

```{r eval=FALSE}
str(mort)
summary(mort)

str(pop)
summary(pop)

str(corr)
summary(corr)

str(env)
summary(env))
```

```{r eval=TRUE, ehco=TRUE}
var_summary <- function(data, var) {
  data %>%
    summarise(n = n(), min = min({{ var }}), max = max({{ var }}))
}
  
  mtcars %>%
  group_by(cyl) %>%
  var_summary(mpg)


```


# Strings and `stringr`

## Combining Strings
```{r strings, echo=TRUE}
string = "This is a string!"
str_c("x", "y")

str_c("x", "y", sep = ", ")

str_c("x", "y", sep = "-")

str_c("prefix-", c("a", "b", "c"), "-suffix")

```


# Strings

## String Length

```{r, echo=TRUE}

str_length(c("a", "R for data science", NA))

```



## Changing Case

There are a few arguments for changing the cases of strings. `str_to_lower()`, `str_to_upper()` or `str_to_title()`

```{r, echo=TRUE}
str_to_lower(x)
str_to_upper(x)
str_to_title(x)

```




## Subsetting Strings

### You can extract parts of a string using `str_sub()`. As well as the string, `str_sub()` takes start and end arguments which give the (inclusive) position of the substring

### negative numbers count backwards from end

```{r subset strings, echo=TRUE}

x <- c("Apple", "Banana", "Pear")
str_sub(x, 1, 3)

str_sub(x, -3, -1)


```

# Regex

### Regexps are a very terse language that allow you to describe patterns in strings. They take a little while to get your head around, but once you understand them, you’ll find them extremely useful.

### To learn regular expressions, use `str_view()` and `str_view_all()`. These functions take a character vector and a regular expression, and show you how they match. We’ll start with very simple regular expressions and then gradually get more and more complicated using `stringr`

```{r, echo=TRUE}
x <- c("apple", "banana", "pear")
str_view(x, "an")

```

# Regex

### The next step up in complexity is `.`, which matches any character (except a newline):

```{r, echo=TRUE}
str_view(x, ".a.")

```

### But if `.` matches any character, how do you match the actual character `.`? You need to use an .yellow[escape] to tell the regular expression you want to match it exactly, not use its special behaviour. Represented by two `\\` back slashes or square brackets `[ ]`.

```{r, echo=TRUE}
str_view(c("abc", "a.c", "bef"), "a\\.c")
str_view(c("abc", "a.c", "bef"), "a[.]c")

```


```{r, echo=TRUE}
str_view(x, "^a")
```

```{r, echo=TRUE}
str_view(x, "a$")
```



## More Examples:

```{r, echo=TRUE}
x <- c("apple pie", "apple", "apple cake")
str_view(x, "apple")

```

### Only match the whole word 

```{r, echo=TRUE}
str_view(x, "^apple$")

```


## str_replace

### Replace all vowels with '-'

```{r, echo=TRUE}
x <- c("apple", "pear", "banana")
str_replace(x, "[aeiou]", "-")
```

For thorough documentation on regex and stringr review the R for Data Science Handbook: https://r4ds.had.co.nz/strings.html


# Lists

### Lists are the next step up from named vectors. Lists can contain other lists

```{r, echo=TRUE}
x <- list(1, 2, 3)
x
```

### We can combine multiple data sets into a list, as we'll see in the exercises

#Lists

```{r}

x1 <- list(c(1, 2), c(3, 4))
x2 <- list(list(1, 2), list(3, 4))
x3 <- list(1, list(2, list(3)))

```



#Purrr

### Works well with lists for applying a function to everything in a given list
### `map` replaces .green[for loops] in code and is succinct and easier to read.

```{r, echo=TRUE}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

```


Say we wanted the `median` of each:

```{r, echo=TRUE}
median(df$a)

median(df$b)

median(df$c)

median(df$d)

```


# Using `map`

```{r, echo=TRUE}
map_dbl(df, median)


```