---
title: "Data Management with R"
author: "Noushin Nabavi"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site

documentclass: book
fontsize: 11pt
geometry: margin=1in

output:
  bookdown::gitbook:
    css: style.css

  bookdown::pdf_book:
    latex_engine: xelatex
    keep_tex: true
classoption: oneside
bibliography: references.bib
link-citations: true
---

# Welcome {-}

This book provides practical teaching materials and examples for **data management and reproducible analysis using R**. The focus is on building clear, structured, and repeatable workflows for working with data in an applied setting.

The materials are designed for learners who are developing confidence with R, RStudio Projects, the tidyverse, R Markdown, and bookdown. The examples use instructional datasets and emphasize the process of preparing, cleaning, joining, reshaping, visualizing, and communicating data.

---

# Overview {-}

This book helps learners:

- Develop practical data-wrangling skills using the tidyverse
- Apply reproducible project structures for consistent analysis
- Work with tidy data principles
- Read, clean, reshape, and join datasets
- Use strings, dates, and regular expressions in data cleaning
- Create clear tables, figures, and analysis outputs
- Communicate results through R Markdown, bookdown, dashboards, and storyboards

The goal is not advanced statistical modelling. Instead, the emphasis is on building reliable workflows that support clear, organized, and reproducible analysis.

---

# How to Use This Book {-}

Each chapter introduces a data management topic and includes worked examples, code chunks, explanations, and practice activities. Learners are encouraged to:

- Read each section carefully
- Run the code chunks in RStudio
- Modify examples to test their understanding
- Keep raw data unchanged
- Save generated outputs in the `Outputs/` folder
- Use clear file names and reproducible project structure

The chapters are intended to be followed in order, but individual chapters can also be used as standalone reference materials.

---

# Recommended Chapter Flow {-}

The book is organized into the following learning sequence:

1. **Project Setup**  
   Set up an RStudio Project, organize folders, and understand reproducible workflows.

2. **Tidyverse Basics**  
   Learn core tidyverse functions for reading, selecting, filtering, mutating, and summarizing data.

3. **Joining Data**  
   Practice combining datasets using keys and joins, with examples from the Titanic dataset.

4. **Data Cleaning**  
   Work with raw files, column names, missing values, reshaping, and structured cleaning steps.

5. **Strings and Regular Expressions**  
   Use `stringr`, `separate()`, `str_extract()`, and regex patterns to clean and parse text data.

6. **Data Visualization**  
   Create clear visual summaries using `ggplot2`, including line plots, bar charts, and grouped visualizations.

7. **Exploratory Data Analysis**  
   Build an organized EDA workflow using summaries, plots, missing-value checks, and data validation.

8. **Communicating Results**  
   Use storyboards, dashboards, and reporting structure to communicate data insights clearly.

---

# Project Structure {-}

A recommended project structure is shown below:

```text
project/
│
├── index.Rmd
├── 01_project_setup.Rmd
├── 02_tidyverse_basics.Rmd
├── 03_joining_data.Rmd
├── 04_data_cleaning.Rmd
├── 05_strings_regex.Rmd
├── 06_visualization.Rmd
├── 07_tidyverse_analysis.Rmd
├── 08_storyboard.Rmd
│
├── _bookdown.yml
├── _output.yml
├── style.css
├── data_management.Rproj
│
├── Raw Data/
├── Outputs/
├── docs/
├── archive/
│
├── README.md
├── LICENSE
└── .gitignore
```

Raw data should remain in `Raw Data/`. Files created by analysis should be saved in `Outputs/`. Rendered book files can be saved in `docs/` if the book is being published with GitHub Pages.

---

# Required R Packages {-}

The book uses several R packages, including:

```{r index-install-packages, eval=FALSE}
install.packages(c(
  "bookdown",
  "tidyverse",
  "ggplot2",
  "readxl",
  "lubridate",
  "stringr",
  "purrr",
  "knitr",
  "kableExtra",
  "here",
  "flexdashboard"
))
```

Load commonly used packages with:

```{r index-load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(readxl)
library(lubridate)
library(stringr)
library(purrr)
```

---

# Building the Book {-}

To build the book from RStudio, open the project file:

```text
data_management.Rproj
```

Then use the **Build Book** option in the Build pane.

You can also render the book from the R console:

```{r index-render-book, eval=FALSE}
bookdown::render_book("index.Rmd")
```

---

# Data and Reproducibility Notes {-}

This project uses relative file paths and a structured folder system. This makes the work easier to share, rerun, and maintain.

Recommended practices:

- Do not edit raw data files directly
- Keep all raw datasets in `Raw Data/`
- Save processed files in `Outputs/`
- Use clear and consistent file names
- Use RStudio Projects to keep paths stable
- Comment code where needed
- Keep analysis steps in a logical order

All datasets used in the course are simulated, fictitious, or instructional only. They are intended for teaching data management concepts, not for drawing real-world research conclusions.

---

# Intended Audience {-}

This book is intended for learners who have basic familiarity with R and RStudio and want to build stronger applied data management skills. It can be used for:

- Self-guided learning
- Training workshops
- Teaching support materials
- Practice exercises
- Templates for reproducible workflows

---

# Licence {-}

For educational use. See `LICENSE` for details.
